A Simple Type Theory for Category Theory

### 2022-04-25 paper/paper.tex
GAT progress: double cats w/ displays, restrictions, cat of elements
GAT Profunctor Multiplicatives
more GATs, start initiality, sketch internal DRA/PRA

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
%% \newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}
\newcommand{\varr}[2]{\text{Fun}{#1}{#2}}
\newcommand{\harr}[2]{\text{Rel}{#1}{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

% Algebraic notation
\newcommand{\algCtx}{\textrm{Ctx}}
\newcommand{\algSubst}{\textrm{Subst}}
\newcommand{\algTy}{\textrm{Type}}
\newcommand{\algTm}{\textrm{Term}}
\newcommand{\algCat}{\textrm{Cat}}
\newcommand{\algVarr}{\textrm{Functor}}
\newcommand{\algHCtx}{\textrm{ProfCtx}}
\newcommand{\algHSubst}{\textrm{ProfSubst}}
\newcommand{\algHarr}{\textrm{Prof}}
\newcommand{\algTrans}{\textrm{Trans}}

@@

@@

@@

    \inferrule
    {}
    {\id_h : \algVarr(C,C)}

    \id_h \circ_{vs} \gamma = \id_h

    (h \circ \id) = h\and \id \circ h = h

@@

@@

    % strict equipment (horizontal)
    % strict equipment  (displays)

    % 2-cells: substitutions
    % vertical category
    % horizontal category
    % interchange law

    % 2-cells: transformations : Q: is any of this necessary or should
    % we just conflate transformations with substitutions into a
    % display?
    % 
    % action of trans on subst with assoc, unit
    % identity trans with unit
    % embedding of transformations into substitutions

    \inferrule
    {P : \algHarr(\Gamma,C',D')
    \and f : \algVarr(\Gamma,C,C')\and
    \and g : \algVarr(\Gamma,D,D')}
    {P \circ_r (f,g) : \algHarr(\Gamma,C,D)}

@@

    \inferrule
    {\psi : \algHSubst(f',g',\Phi,\Psi)\and \phi : \algHSubst(f,g,\Sigma,\Phi)
    }
    {\psi \circ \phi : \algHSubst(f' \circ f, g' \circ g, \Sigma, \Psi)}

    \inferrule
    {}
    {restrict(f,g,\phi) : \algHSubst(\id,\id,\Psi,\Phi)}
  \end{mathpar}
  \caption{Equipment with Displays}
  \label{fig:alg-dc-disp}
\end{figure}

\begin{figure}
  \begin{mathpar}
    % Cat Multiplicatives
    % Cat of Elements
    % Presheaf cats

    % Prof Multiplicatives
    % Unit
    % Tensor
    % Hom

    % Internalization
    % Cats
    % Functor
    % Prof
    % ProfCtx?
    % Trans

@@

    % Prof Additives
    % Product
    % Terminal
  \end{mathpar}
  \caption{UMPs in an Equipment}
\end{figure}

>>>>

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}
\newcommand{\varr}[2]{\text{Fun}{#1}{#2}}
\newcommand{\harr}[2]{\text{Rel}{#1}{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

% Algebraic notation
\newcommand{\algCtx}{\textrm{Ctx}}
\newcommand{\algSubst}{\textrm{Subst}}
\newcommand{\algTy}{\textrm{Type}}
\newcommand{\algTm}{\textrm{Term}}
\newcommand{\algCat}{\textrm{Cat}}
\newcommand{\algVarr}{\textrm{Functor}}
\newcommand{\algHCtx}{\textrm{ProfCtx}}
\newcommand{\algHSubst}{\textrm{ProfSubst}}
\newcommand{\algHarr}{\textrm{Prof}}
\newcommand{\algTrans}{\textrm{Trans}}
\newcommand{\algElts}{\textrm{Elts}}
\newcommand{\algEltsI}{\textrm{EltsI}}

\newcommand{\algLHom}{\textrm{LHom}}
\newcommand{\algLHomI}{\textrm{LHomI}}
\newcommand{\algLHomE}{\textrm{LHomE}}
\newcommand{\algRHom}{\textrm{RHom}}
\newcommand{\algRHomI}{\textrm{RHomI}}
\newcommand{\algRHomE}{\textrm{RHomE}}

\newcommand{\algUnit}{\textrm{Unit}}
\newcommand{\algUnitI}{\textrm{UnitI}}
\newcommand{\algUnitE}{\textrm{UnitE}}

\newcommand{\algTensor}{\textrm{Tensor}}
\newcommand{\algTensorI}{\textrm{TensorI}}
\newcommand{\algTensorE}{\textrm{TensorE}}

\newcommand{\algCatTy}{\textrm{Cat}}
\newcommand{\algCatQt}{\textrm{CatQt}}
\newcommand{\algCatUnqt}{\textrm{CatUnQt}}

\newcommand{\algTransTy}{\textrm{Trans}}
\newcommand{\algTransQt}{\textrm{TransQt}}
\newcommand{\algTransUnqt}{\textrm{TransUnQt}}

\newcommand{\algVarrQt}{\textrm{FunctQt}}
\newcommand{\algVarrUnqt}{\textrm{FunctUnQt}}

@@

%% Can we do pras?
%%
%% A PRA mode theory is given by a free 2-category
%% 1. Each mode m is a CwF whose context cat is [m]
%% 2. Each 1-cell μ : m -> n is a functor [μ] : [n] -> [m] that has a DRA and is a PRA
%% 3. Each 2-cell α : μ => ν is a 2-cell

%% Let C, D be CwFs.
%% A DRA is a pair of
%% 1. A functor  L : C -> D
%% 2. A morphism R : L^*(D_Ty) -> C_ty in ∫C, i.e., a morphism taking LΓ |- A to Γ |- RA
%% 3. An iso     I : L^*(D_tm)?? =~ R^*(C_tm) in ∫(L^*(DTy)) a morphism taking LΓ |- M : A to Γ |- I(M) : RA

%% Let C, D such that C has a terminal object.
%% G : C -> D is a PRA if
%% G / 1 : C -> D / G1 is a right adjoint

%% What about fixed points?
%% Say C has a terminal object
%% fix : Hom(a,a) -> Hom(1, a)
%% satisfying 
%% fix[f] = f o fix[f]?
%% We can write down fix, but not the equation I think

@@

TODO: include all of the interactions with substitutions and restrictions.

@@

    \inferrule
    {}
    {\id_v : \algVarr(C,C)}

    \id_v \circ_{vs} \gamma = \id_h

    (f \circ_v \id_v) = f\and \id_v \circ_v f = f

@@

    \inferrule
    {\Psi : \algHCtx(D,E) \and \algHCtx(C,D)}
    {\Psi \circ_h \Phi : \algHCtx(C,E)}

    (\Psi \circ_h \Phi) \circ_{hs} \gamma = (\Psi \circ_{hs} \gamma) \circ_h (\Phi \circ_s \gamma)

    (\Sigma \circ_h \Psi) \circ_h \Phi = \Sigma \circ_h (\Psi \circ_h \Phi)

    \inferrule
    {}
    {\id_h : \algHCtx(C,C)}

    \id_h \circ_{hs} \gamma = \id_h

    (\Phi \circ_h \id_h) = \Phi\and \id_h \circ_h \Phi = \Phi

@@

    \inferrule{P : \algHarr(C,D)}{i(P) : \algHCtx(C,D)}

    % 2-cells: substitutions
    % vertical category
    % horizontal category
    % interchange law
    \inferrule
    {\psi : \algHSubst(\Phi, f',g',\Psi)\and \phi : \algHSubst(\Sigma,f,g,\Phi)
    }
    {\psi \circ_{2v} \phi : \algHSubst(\Sigma, f' \circ f, g' \circ g, \Psi)}

    \inferrule
    {}
    {\id_{2v} : \algHSubst(\Phi,\id,\id,\Phi)}

    (\phi \circ_{2v} \psi) \circ_{2v} \sigma = \phi \circ_{2v} (\psi \circ_{2v} \sigma)\\
    \phi \circ_{2v} \id_{2v} = \phi \and \phi = \id_{2v} \circ_{2v} \phi\\

    \inferrule
    {\phi_1 : \algHSubst(\Psi_1,f,g,\Phi_1)\and \phi_2 : \algHSubst(\Psi_2,f,g,\Phi_2)
    }
    {\phi_1 \circ_{2h} \phi_2 : \algHSubst(\Psi_1 \circ_h \Psi_2, f, h, \Phi_1 \circ_h \Phi_2)}

    \inferrule
    {}
    {\id_{2h} : \algHSubst(\id_h,f,f,\id_h)}

    (\phi \circ_{2h} \psi) \circ_{2h} \sigma = \phi \circ_{2h} (\psi \circ_{2h} \sigma)\\
    \phi \circ_{2h} \id_{2h} = \phi \and \id_{2h} \circ_{2h} \phi = \phi\\

    (\phi_1 \circ_{2h} \phi_2) \circ_{2v} (\psi_1 \circ_{2h} \psi_2) = (\phi_1 \circ_{2v} \psi_1) \circ_{2h} (\phi_2 \circ_{2v} \psi_2)

  \end{mathpar}
  \caption{Double Category with Display Pro-arrows}
  \label{fig:alg-dc-disp}
\end{figure}

\begin{figure}
  \begin{mathpar}
    % Restrictions
    \inferrule
    {P : \algHarr(\Gamma,C',D')
    \and f : \algVarr(\Gamma,C,C')\and
    \and g : \algVarr(\Gamma,D,D')}
    {P \circ_r (f,g) : \algHarr(\Gamma,C,D)}

@@

    % Prof Multiplicatives
    % LHom
    \inferrule
    {P : \algHarr(C, D) \and Q : \algHarr(C,E)}
    {\algLHom(P,Q) : \algHarr(D,E)}

    \algLHom(P,Q) \circ_{hs} \gamma = \algLHom(P \circ_{hs} \gamma, Q \circ_{hs} \gamma)\and
    \algLHom(P,Q) \circ_{r} (f,g) = \algLHom(P \circ_r (\id,f), Q \circ_r(\id,g))

    \inferrule
    {}
    {\algLHomE(P,Q) : \algHSubst(i(P)\circ_h i(\algLHom(P,Q)),\id_v,\id_v,i(Q))}

    \inferrule
    {t : \algHSubst(i(P) \circ_h \Phi, \id,g, i(Q))}
    {\algLHomI(\Phi,g,t,P,Q) : \algHSubst(\Phi,\id,g,i(\algLHom(P,Q)))}

    \algLHomE \circ_{2v} (\id_v(P) \circ_h \algLHomI(t)) = t\and
    \theta = \algLHomI(\algLHomE \circ_{2v} (\id_v(P) \circ_h \theta))\and

    % RHom
    \inferrule
    {P : \algHarr(C, E) \and Q : \algHarr(D,E)}
    {\algRHom(P,Q) : \algHarr(C, D)}

    \algRHom(P,Q) \circ_{hs} \gamma = \algRHom(P \circ_{hs} \gamma, Q \circ_{hs} \gamma)\and
    \algRHom(P,Q) \circ_{r} (f,g) = \algRHom(P \circ_r (f, \id), Q \circ_r(g,\id))

    \inferrule
    {}
    {\algRHomE(P,Q) : \algHSubst(i(\algRHom(P,Q)) \id_h i(Q),\id_v,\id_v,i(P))}

    \inferrule
    {t : \algHSubst(\Phi \circ_h i(Q), f, \id, i(P))}
    {\algRHomI(\Phi,f,t,P,Q) : \algHSubst(\Phi,\id,g,i(\algLHom(P,Q)))}

    \algRHomE \circ_{2v} (\algRHomI(t) \circ_h \id(Q)) = t\and
    \theta = \algRHomI(\algRHomE \circ_{2v} (\theta \circ_h \id(Q)))\and

    % Unit
    \inferrule
    {C : \algCat}
    {\algUnit(C) : \algHarr(C,C)}

    \algUnit(C) \circ_{hs} \gamma = \algUnit(C \circ_{cs} \gamma)

    \inferrule
    {}
    {\algUnitI(C) : \algSubst(\id_h(C),\id,\id,i(\algUnit(C)))}

    \inferrule
    {t : \algHSubst(\Phi \id_h \Psi,f,g,i(R))}
    {\algUnitE(t) : \algHSubst((\Phi \circ_h i(\algUnit(C)) \circ_h \Psi), f, g, i(R))}

    \algUnitE(t) \circ_{2v} (\id_{2v}(\Phi) \circ_h \algUnitI(C) \circ_h \id_{2v}(\Psi)) = t\and
    s = \algUnitE(s \circ_{2v} (\id_{2v}(\Phi) \circ_h \algUnitI(C) \circ_h \id_{2v}(\Psi)))

    % Tensor
    \inferrule
    {P : \algHarr(C,D) \and Q : \algHarr(D,E)}
    {\algTensor(P,Q) : \algHarr(C,E)}

    \algTensor(P,Q) \circ_{hs} \gamma = \algTensor(P \circ_{hs} \gamma, Q \circ_{hs} \gamma)\and
    \algTensor(P,Q) \circ_{r} (f,g) = \algTensor(P \circ_r (f,\id), Q \circ_r (\id,g))

    \algTensorI(P,Q) : \algHSubst(i(P) \circ_h i(Q),\id,\id,i(\algTensor(P,Q)))

    \inferrule
    {t : \algHSubst(\Phi \circ_h i(P) \circ_h i(Q) \circ_h \Psi, f,g,i(R))}
    {\algTensorE(t) : \algHSubst(\Phi \circ_h i(\algTensor(P,Q)) \circ_h \Psi, f,g,i(R))}

    \algTensorE(t) \circ_{2v} (\id_{2v}(\Phi) \circ_{2h} \algTensorI(P,Q) \circ_{2h} \id_{2v}(\Psi)) = t\and
    s = \algTensorE(s \circ_{2v} (\id_{2v}(\Phi) \circ_{2h} \algTensorI(P,Q) \circ_{2h} \id_{2v}(\Psi)))

    % Cat Multiplicatives
    % Cat of Elements
    \inferrule
    {P : \algHarr(\Gamma,C,D)}
    {\algElts(C,D,P) : \algCat(\Gamma)}

    \inferrule
    {}
    {\pi^- : \algVarr(\algElts(C,D,P),C)}

    \inferrule
    {}
    {\pi^+ : \algVarr(\algElts(C,D,P),D)}

    \inferrule
    {}
    {\pi^p : \algHSubst(\id_h(\algElts(C,D,P)),\pi^-,\pi^+,P)}

    \inferrule
    {f : \algVarr(E,C) \and g : \algVarr(E,D)\and
      t : \algHSubst(\id_h(E), f, g, P)
    }
    {\algEltsI(t,f,g) : \algVarr(E,\algElts(C,D,P))}

    \pi^- \circ_v \algEltsI(t,f,g) = f\and
    \pi^+ \circ_v \algEltsI(t,f,g) = g\and
    \pi^p \circ_{2v} \algEltsI(t,f,g) = \alpha

    \inferrule
    {h : \algVarr(E,\algElts(C,D,P))}
    {h = \algEltsI(\pi^-\circ_v h,\pi^+\circ_vh,\pi^p \circ_{2v} \id_{2h}(h))}

    % Presheaf cats

@@

    % Prof Additives
    % Product
    % Terminal
  \end{mathpar}
  \caption{Multiplicatives}
\end{figure}

\begin{figure}
  \begin{mathpar}
    % Internalization
    % Cats
    \inferrule
    {}
    {\algCatTy : \algTy(1)}

    \inferrule
    {C : \algCat(\Gamma)}
    {\algCatQt(C) : \algTm(\Gamma,\algCatTy \circ !)}

    \algCatQt(C) \circ_{tms} \gamma = \algCatQt(C \circ_{cs} \gamma)\and

    \inferrule
    {M : \algTm(\Gamma,\algCatTy \circ !)}
    {\algCatUnqt(M) : \algCat(\Gamma)}

    \algCatUnqt(M) \circ_{cs} \gamma = \algCatUnqt(M \circ_{tms} \gamma)\and
    \algCatUnqt(\algCatQt(C)) = C\and
    M = \algCatQt(\algCatUnqt(M))

    % Functor
    % Prof
    % ProfCtx?
    % Trans
    \inferrule
    {P : \algHarr(\Gamma,C,C)}
    {\algTransTy(P) : \algTy(\Gamma)}

    \inferrule
    {t : \algHSubst(\id_h,\id_v,\id_v,i(P))}
    {\algTransQt(t) : \algTm(\Gamma,\algTransTy(P))}

    \algTransQt(t) \circ_{tms} \gamma = \algTransQt(t \circ_{2s} \gamma)

    \inferrule
    {}
    {\algTransUnqt : \algHSubst(ext(\Gamma,\algTransTy(P)),\id_h,\id_v,\id_v,i(P \circ_{hs} \pi_1))}

    \algTransUnqt \circ_{2s} (\gamma,\algTransQt(t)) = t\\

    % M : Tm(G,∀ α. P α α)
    % unquote : Subst(G,t : ∀ α:C. P α α; α:C, id,id, P α α)
    % M = unquote()
    M = \algTransQt(\algTransUnqt \circ_{2s} ext_s(\id,M))
  \end{mathpar}
  \caption{Internalization}
\end{figure}

\subsection{Equivalence between Natural Deduction and Combinatory Syntaxes}

Now we aim to prove that our natural deduction syntax of SCT is sound
and complete for the algebraic notion of hyperdoctines of double
categories with displays.
%
In category-theoretic terms, we want to show that the syntax of SCT
presents an \emph{initial object} in the category of models.
%
This is proved in two steps: first, we show that SCT is a model

\begin{construction}{Completeness}
  SCT presents a hyperdoctrine of double categories with displays.
\end{construction}
\begin{proof}
  First, the sorts.
  \begin{align*}
    \eta(\algCtx) &= \{ \Gamma | \Gamma \isadtctx \}\\
    \eta(\algSubst(\Delta,\Gamma)) &= \{ \gamma | \eta(\Delta) \vdash \gamma : \eta(\Gamma) \}\\
    \eta(\algTy(\Gamma)) &= \{ A |  \eta(\Gamma) \vdash A \isaTy \}\\
    \eta(\algTm(\Gamma,A)) &= \{ M | \eta(\Gamma) \vdash M : \eta(A) \}\\
    \eta(\algCat(\Gamma)) &= \{ \cat C | \eta(\Gamma) \vdash \cat C \isaCat \}\\
    \eta(\algVarr(\Gamma,C,D)) &= \{ a | \eta(\Gamma)|\alpha:\eta(C) \vdash a : \eta(D) \}\\
    \eta(\algHarr(\Gamma,C,D)) &= \{ R | \eta(\Gamma)|\alpha:\eta(C);\beta:\eta(D) \vdash R \isaSet \}\\
    \eta(\algHCtx(\Gamma,C,D)) &= \{ \Phi | \eta(\Gamma) \vdash \Phi \isavectx \wedge d^-(\Phi) = \eta(C) \and d^+(\Phi) = \eta(D) \}\\
    \eta(\algHSubst(\Gamma, \Psi, f, g, \Phi)) &= \{ \phi | \eta(\gamma)\pipe \eta(\Psi) \vdash \phi : \eta(\Phi) \wedge d^-\phi = \eta(f) \wedge d^+\phi = \eta(g) \}\\
    1 &= \cdot\\
    ! &= \cdot\\
    ext(\Gamma,A) = \Gamma, x:A \\
    \cdots\\
    \Phi \circ_h \Psi &= \Phi \jnctx \Psi\\
    \id_h(\cat C) &= \alpha:\cat C\\
    i(\cat C, \cat D, R) &= \alpha:\cat C, x: R, \beta:\cat D\\
    \phi \circ_{2v} \psi &= \phi[\psi]\\
    \id_{2v}(\Phi) &= \id_\Phi\\
    \id_{2h}(\alpha:\cat C \vdash a : \cat D) &= a/\beta\\
    \phi \circ_{2h} \psi &= \phi \jnctx \psi
  \end{align*}
\end{proof}
\begin{corollary}
  There is a unique homomorphism of HDCDs
  \[ \eta : Comb \to SCT \]
\end{corollary}

\begin{construction}{Soundness}
  We construct a homomorphism of HDCDs
  \[ i : SCT \to Comb \]
  such that the composite
  \[ \eta \circ i : SCT \to SCT \]
  is equal to the identity.
\end{construction}
\begin{proof}
  \begin{align*}
    i(\cdot) &= 1\\
    i(\Gamma,x:A)  &= ext(i(\Gamma),ext(A))\\
    \cdots
    i(\Gamma \pipe \alpha:\cat C \vdash \alpha:\cat C) &= \id_v\\
    i(\Gamma \pipe \alpha:\cat C \vdash M a :\cat C) &= (\algVarrUnqt \circ_{vs} (\id, (i(M)))) \circ_v i(a)\\
    % Presheaf intro
    % elements intro
    i(\Gamma \pipe \alpha:\cat C \vdash (a,t,b) : \Sigma_{\alpha:\cat C;\beta:\cat D} R) &= \algEltsI(i(a),i(t),i(b))\\
    %
  \end{align*}
\end{proof}

\begin{corollary}
  SCT is initial in the category of HDCDs.
\end{corollary}
\begin{proof}
  Since Comb is initial, the composite $i \circ \eta$ is the identity,
  and as above the composite $\eta \circ i$ so SCT is isomorphic to
  Comb, and therefore it is also initial.
\end{proof}
